---
title: "CCSM CDS Analytics"
date: today
date-format: D MMMM YYYY
format: html
editor: source
author: Michael Nosal (mnosal@mitre.org)
version: 0.1.1
---

```{r setup, include=FALSE}
# setup for data processing
knitr::opts_chunk$set(echo = FALSE, message = FALSE)

source('R/_load_libraries.R')
source('R/_utility_functions.R')
base::readRenviron("_environment")
base::readRenviron("_environment.local")
base::readRenviron("_environment.dev")
```

`r as.Date(now(), format = "%Y-%m-%d")`

©Copyright 2025 - MITRE Corporation. For limited distribution.

# Cervical Cancer Screening and Management (CCSM) Clinical Decision Support (CDS) Pilot Analytics

## Analysis through logging CDS patient expressions

When the CCSM CDS processes a patient, it automatically evaluates over 800 expressions about the patient history. These items define attributes about the patient which are then used to evaluate the logical statements which represent the clinical guideline recommendations.


## Log Entries
The CCSM CDS Dashboard is set to log patient information every time it 'fires' the CDS and generates a recommendation. This occurs when the Dashboard is launched for a patient, when the user hits 'refresh', and every time a toggle switch is changed. This means the Dashboard may generate multiple log entries during the course of an 'encounter' with the patient. A patient may also be viewed in the Dashboard on multiple occasions, generating log entries each time. 

```{r logfile_parsing}
# Load available logfiles

# Default is to look in ./data/logfiles
datafiles_path <- here('data','logfiles')

# LOGFILE_DIRECTORY environment variable may point to a different location for logfiles
logfiles_env_directory <- Sys.getenv("LOGFILES_DIRECTORY")
logfiles_start_date <- as.Date(Sys.getenv("LOGFILES_START_DATE"))
logfiles_end_date <- as.Date(Sys.getenv("LOGFILES_END_DATE"))

logfiles_date_range <- paste0("ccsmcds-",seq(logfiles_start_date, logfiles_end_date, "days"))

logfiles_directory <- ifelse(
  logfiles_env_directory != "",
  logfiles_env_directory,
  datafiles_path)

# Read all .log and .log.gz files in the directory
all_logfiles <- dir(logfiles_directory, pattern = "*.log|*.log.gz")
# now check if there is a date range to compare against
if (length(logfiles_date_range) > 0) {
    matching_logfiles <- all_logfiles |> keep(\(x) any(str_detect(logfiles_date_range,strsplit(x,".",fixed=TRUE)[[1]][1])))
} else {
  matching_logfiles <- all_logfiles
}

logfile_count <- list(
  logfiles = 0,
  total_entries = 0,
  unique_patients = 0
)
logfile_count$logfiles <- length(matching_logfiles)

# Stop if there are no files to read length(files) == 0
if (logfile_count$logfiles == 0) {
  print(paste("No matching log files found at", logfiles_directory))
  knitr::knit_exit()
} else {

# load individual JSON files into single unexpanded tbl_json
# Logfiles are in jsonl (JSON Lines) format, either compressed (.log.gz) or
# uncompressed (.log)
logdata <- matching_logfiles %>%
       map_df(~read_json(file.path(logfiles_directory, .), format="jsonl"))

logfile_count$total_entries <- nrow(logdata)

# extract just the id and timestamps as a data frame
logdata_timestamps <- as_tibble(
  logdata %>%
    spread_all(recursive = FALSE) %>%
    mutate(timestamp = ymd_hms(timestamp), id= as_factor(id)) %>%
    select(id,timestamp))

logdata_start_date <- as.Date(min(logdata_timestamps$timestamp))
logdata_end_date   <- as.Date(max(logdata_timestamps$timestamp))

# Now we can produce a timestamp scatterplot to show when data from the CDS was logged
# logdata_timestamps

# df <- data.frame(
#   dates = as.POSIXct(date(logdata_timestamps$timestamp)),
#   times = as.POSIXct(logdata_timestamps$timestamp)
# )
# ggplot(df, aes(x=dates, y=times)) +
#   geom_point() +
#   scale_y_datetime(breaks=date_breaks("4 hour"), labels=date_format("%H:%M")) +
#   scale_x_datetime(breaks = date_breaks("1 day")) +
#   theme(axis.text.x=element_text(angle=90))

# Perform a sanity check on the logdata table itself
# The top level data is from the logger service (id, level, message, service, timestamp)
# This should be a 5x3 tibble, with the count column n the same for all and equal to the number of logfile_total_entries.
# If not, there is something unusual in the log data.
top_level_log_types <- logdata %>% gather_object %>% json_types %>% count(name, type)

# If nrow(top_level_log_types) != 5, something is wrong
# If length(unique(top_level_log_types)) != 1, something is wrong
# If unique(top_level_log_types$n) != nrow(logdata), something is wrong
# unique(top_level_log_types$level) == "info"
# 

# Extract the message object
# "message": {
#   "cdsApplyEnd": 1745520994391,
#   "cdsApplyStart": 1745520986793,
#   "patientInfo": { OBJ },
#   "patientReference": "Patient/123abc456def",
#   "payload": { OBJ },
#   "timeRequestSent": "2025-04-24T18:56:34.391Z",
#   "toggleStatus": { OBJ }
# }
# The message object is from the Dashboard, and is where Dashboard parameters are sent.

logdata_messages <- as_tibble(
  logdata %>% 
    spread_values(id = jstring(id)) %>% 
    enter_object(message) %>%
    spread_all(recursive = FALSE) %>%
    enter_object(toggleStatus) %>%
    spread_all %>%
    select(-document.id) %>%
    mutate(cdsApplyTime = cdsApplyEnd - cdsApplyStart,
           patientReference = as_factor(patientReference),
           timeRequestSent = ymd_hms(timeRequestSent))
)

# At this point, we have a tibble of 
# cdsApplyStart, cdsApplyEnd, patientReference, timeRequestSent and the 5 toggleStatus values
# isImmunosuppressed, isPregnant, isPregnantConcerned, isSymptomatic and isToggleChanged
# we'll mutate a cdsApplyTime column to be the diff between cdsApplyEnd and cdsApplyStart in milliseconds

# here we can get the number of unique patients in the log by checking patientReference, which is the patient's FHIR identifier
logfile_count$unique_patients <- length(unique(logdata_messages$patientReference))
 
# Count the number of log entries for every patient and show the top 10 by number of log entries
# to get a sense of how often the CDS was triggered on patients
# X patients had 23 log entries
# Y patients had 21 log entries
# etc, etc
patient_message_summary <- logdata_messages %>%
  group_by(patientReference) %>%
  summarize(count = n())

# We can also use this table to get a summary of how many times each toggle switch was used
# returns a named vector, use toggle_summary["isPregnant"] to access named value
toggle_summary <- colSums(logdata_messages %>% select(isImmunosuppressed, isPregnant, isPregnantConcerned, isSymptomatic, isToggleChanged))

# Produce table showing how many times each toggle switch was used
}
```
## Logfile Data Summary

This report evaluated a total of `r logfile_count$logfiles` log files from `r logdata_start_date` to `r logdata_end_date`. These log files contained a total of `r logfile_count$total_entries` total entries. There were `r logfile_count$unique_patients` unique patient identifiers in these entries. `r matching_logfiles`

```{r parse-libraries}
logdata_dashboard <-tidyjson::as_tibble(
  logdata %>%
    spread_values(id = jstring(id)) %>% 
    enter_object(message) %>%
      enter_object(payload,DashboardLibrary) %>%
        spread_all
)

logdata_management_library <-tidyjson::as_tibble(
  logdata %>%
    spread_values(id = jstring(id)) %>% 
    enter_object(message) %>%
      enter_object(payload,ManagementLibrary) %>%
        spread_all(recursive = FALSE)
)

logdata_common <-tidyjson::as_tibble(
  logdata %>%
    spread_values(id = jstring(id)) %>% 
    enter_object(message) %>%
      enter_object(payload,ManageCommonAbnormality) %>%
        spread_all(recursive = FALSE) %>%
        mutate(DateOfMostRecentHpvReport = date(DateOfMostRecentHpvReport),
               DateOfMostRecentPositiveScreeningResult = date(DateOfMostRecentPositiveScreeningResult))
)

# In the Risk Tables, there are management recommendations which are not associated with specific risk estimates.
# For coding purposes, these are represented as -1 in the lookup tables used by the CDS. We'll need to replace these with "NA"
logdata_relevant_row <-tidyjson::as_tibble(
  logdata %>%
    spread_values(id = jstring(id)) %>% 
    enter_object(message) %>%
      enter_object(payload,ManageCommonAbnormality,RelevantRow) %>%
        spread_all(recursive = FALSE) %>%
        mutate(riskNow = na_if(riskNow, -1), risk5yr = na_if(risk5yr, -1))
)

logdata_rare_abnormality <-tidyjson::as_tibble(
  logdata %>%
    spread_values(id = jstring(id)) %>% 
    enter_object(message) %>%
      enter_object(payload,ManageRareAbnormality) %>%
        spread_all(recursive = FALSE)
)

logdata_special_population <-tidyjson::as_tibble(
  logdata %>%
    spread_values(id = jstring(id)) %>% 
    enter_object(message) %>%
      enter_object(payload,ManageSpecialPopulation) %>%
        spread_all(recursive = FALSE)
)
logdata_collate <-tidyjson::as_tibble(
  logdata %>%
    spread_values(id = jstring(id)) %>% 
    enter_object(message) %>%
      enter_object(payload,CollateManagementData) %>%
        spread_all(recursive = FALSE) %>%
        mutate(MostRecentBiopsyReportDate = date(MostRecentBiopsyReportDate),
               MostRecentCytologyReportDate = date(MostRecentCytologyReportDate),
               DateOfFirstBiopsyAfterMostRecentHpvReport = date(DateOfFirstBiopsyAfterMostRecentHpvReport),
               DateOfMostRecentCytologyBeforeBiopsy = date(DateOfMostRecentCytologyBeforeBiopsy))
)
```

## Patient Demographic Information
```{r patient_demographics}

# Now we will drill into the patientInfo. This contains the relevant information from the Patient resourceType
# Note that 'race' is actually concatenated with ethnicity e.g. "White or Caucasian; Not Hispanic or Latino"
# but is separated by semi-colon, so we can split
# This should include patientReference and filtered to unique patients, so we aren't counting the 
# same patient multiple times
# patientInfo should include:
# patientInfo: {
#    name: 'Marge Simpson',
#    id: [ OBJ ],
#    isPregnant: false,
#    dateOfBirth: { value: [Object] },
#    sexAtBirth: 'Female',
#    age: 64,
#    gender: 'female',
#    primaryLanguage: 'English',
#    race: 'White or Caucasian; zzNot Hispanic or Latino'
# We are mostly interested in age, race, ethnicity, pregnancy status at the moment
# NOTE: Ethnicity is separated by "; " 
# Table arranged by timeRequestSent Descending
logdata_all_patientinfo <- tidyjson::as_tibble(
  logdata %>%
    spread_values(id = jstring(id)) %>% 
    enter_object(message) %>%
      spread_values(patientReference = jstring('patientReference'), timeRequestSent = jstring('timeRequestSent')) %>%
      enter_object(patientInfo) %>%
        spread_values(age=jinteger('age'),
                      racestr=jstring('race'),
                      isPregnant = jlogical('isPregnant')) %>%
      select(-document.id, age, racestr, isPregnant) %>%
      mutate(patientReference = as.factor(patientReference),
             timeRequestSent = ymd_hms(timeRequestSent)) %>%
      arrange(desc(timeRequestSent)) %>%
      separate(racestr, c("race","ethnicity"), sep="; ", extra = "drop")
)[,c(1,2,3,4,5,6,7)]

# get the unique patient info for the most recent entry 
logdata_unique_patientinfo <- logdata_all_patientinfo %>% distinct(patientReference, .keep_all = TRUE)

patient_demographics <- logdata_unique_patientinfo %>%
  select(age, race, ethnicity, isPregnant) %>%
  mutate(age_group = cut(age, breaks=c(0,19,24,29,65,120)))

patient_age_groups <- hist(patient_demographics$age, breaks=c(0,19,24,29,65,120), plot = FALSE)

race_list <- tibble(
  race = c("White or Caucasian",
  "Black or African American",
  "Asian",
  "American Indian or Alaska Native",
  "Native Hawaiian or Other Pacific Islander",
  "Other",
  NA)
)

ethnicity_list <- tibble(
  ethnicity = c("Hispanic or Latino","Not Hispanic or Latino",NA)
)

# Note this does not correctly handle race values not included in race_list (e.g. "White" instead of "White or Caucasian")
patient_race_groups <- race_list %>%
  inner_join(patient_demographics) %>%
  group_by(race) %>%
  summarize(count = n()) %>%
  right_join(race_list) %>%
  mutate(count = replace_na(count, 0))

patient_ethnicity_groups <- ethnicity_list %>%
  inner_join(patient_demographics) %>%
  group_by(ethnicity) %>%
  summarize(count = n()) %>%
  right_join(ethnicity_list) %>%
  mutate(count = replace_na(count, 0))

demographics_tbl <- tibble(
  col = c("Total Unique Patients","Age < 20","Age 20-24","Age 25-29","Age 30-65","Age 65+","White or Caucasian","Black or African American","Asian","American Indian or Alaska Native","Native Hawaiian or Other Pacific Islander","Other/Unknown","Hispanic or Latino","Non-Hispanic","Other/Unknown"),
  patients = c(
    logfile_count$unique_patients,
    patient_age_groups$counts[1], # under 20
    patient_age_groups$counts[2], # 20-24
    patient_age_groups$counts[3], # 25-29
    patient_age_groups$counts[4], # 30-65
    patient_age_groups$counts[5], # 65-120
    patient_race_groups %>% filter(race == 'White or Caucasian') %>% .$count,
    patient_race_groups %>% filter(race == 'Black or African American') %>% .$count,
    patient_race_groups %>% filter(race == 'Asian') %>% .$count,
    patient_race_groups %>% filter(race == 'American Indian or Alaska Native') %>% .$count,
    patient_race_groups %>% filter(race == 'Native Hawaiian or Other Pacific Islander') %>% .$count,
    (patient_race_groups %>% filter(is.na(race)) %>% .$count +
      patient_race_groups %>% filter(race == 'Other') %>% .$count),
    patient_ethnicity_groups %>% filter(ethnicity == 'Hispanic or Latino') %>% .$count,
    patient_ethnicity_groups %>% filter(ethnicity == 'Not Hispanic or Latino') %>% .$count,
    patient_ethnicity_groups %>% filter(is.na(ethnicity)) %>% .$count
  ),
  pct = round((patients/logfile_count$unique_patients) * 100,1)
)

kable(demographics_tbl,  col.names = c("Category","Count","%"), align='l',caption="<h1>Demographic Summary</h1>") %>%
  kable_styling(bootstrap_options = c("striped"), full_width = T) %>%
  pack_rows("Age Group (years)", 2, 6) %>%
  pack_rows("Race", 7, 12) %>%
  pack_rows("Ethnicity",13,15)


```

# Patient Pregnancy

The CCSM CDS does attempt to retrieve pregnancy status of a patient, however this is not always well-documented in the patient record. The CDS Dashboard includes two toggle switches related to pregnancy: Patient is Pregnant and Patient has Future Pregnancy Concerns.

We can examine the logged data to determine:

- How many patients were noted as pregnant during any log entry?
- How many patients were noted as pregnant at their most recent 'visit' (the most recent time they were evaluated by the CCSM CDS Dashboard)?
- How many patients who were **not** noted as pregnant in their record, but who used the "Pregnant" toggle at least one time?
- How many patients who were **not** noted as pregnant in their record, but who used the "Future pregnancy concerns" toggle at least one time?

These values are based on unique patients, not the number of log entries for those patients. 

```{r patient_pregnancy}
# This is a little more complicated, as pregnancy status can change between visits
# What we would like:
# How many patients were noted as pregnant during ANY log entry
# How many patients were noted as pregnant at their MOST RECENT visit?
# How many patients were NOT noted as pregnant during their visit but had the isPregnant toggle on?
# How many patients were NOT noted as pregnant during their visit but had the isPregnantConcerned toggle on?
# We can return to this table later to compare against specific histories (e.g. HPV+)

patient_count <- list(
  pregnant_any = 0,
  pregnant_most_recent = 0,
  pregnant_toggle = 0,
  pregnant_concerned_toggle = 0
)

logdata_pregnant_info <- tidyjson::as_tibble(
  logdata %>%
    spread_values(id = jstring(id)) %>%  
    enter_object(message) %>%
      spread_all(recursive = FALSE) %>%
      enter_object(toggleStatus) %>%
      spread_all() %>%
      select(id, isPregnant, isPregnantConcerned) %>%
      rename(isPregnantToggle = isPregnant, isPregnantConcernedToggle = isPregnantConcerned) %>%
      left_join(logdata_all_patientinfo, by="id") %>%
      select( -race, -ethnicity)
)

# Find patients who were noted as pregnant at any logged visit
pregnant_any_time <- logdata_pregnant_info %>% group_by(patientReference) %>%
  filter(isPregnant == TRUE) %>%
  summarize(count = n())
# There will be one row for every patient that was pregnant at any time
# The count column will show the total number of log entries, not encounters or visits
patient_count$pregnant_any <- nrow(pregnant_any_time)

# How many patients were noted as pregnant at their MOST RECENT visit?
# Any visit will correspond to at least on log entry, so the most recent log entry for the patient will do
pregnant_most_recent <- logdata_pregnant_info %>% 
  group_by(patientReference) %>%
  slice_max(timeRequestSent) %>%
  filter(isPregnant == TRUE)
# There will be one row for every patient who was pregnant at their last visit
patient_count$pregnant_most_recent <- nrow(pregnant_most_recent)

# How many patients were NOT noted as pregnant during their visit but had the isPregnant toggle on?
not_pregnant_with_toggle <- logdata_pregnant_info %>%
    filter(isPregnant == FALSE, isPregnantToggle == TRUE) %>%
  group_by(patientReference) %>%
  summarize(count = n())
# there will be one row for every patient who was not pregnant but had the isPregnant toggle on at some point
patient_count$pregnant_toggle <- nrow(not_pregnant_with_toggle)

# How many patients were NOT noted as pregnant during their visit but had the isPregnantConcerned toggle on?
not_pregnant_with_concerned_toggle <- logdata_pregnant_info %>%
    filter(isPregnant == FALSE, isPregnantConcernedToggle == TRUE) %>%
  group_by(patientReference) %>%
  summarize(count = n())
# there will be one row for every patient who was not pregnant but had the isPregnant toggle on at some point
patient_count$pregnant_concerned_toggle <- nrow(not_pregnant_with_toggle)

pregnant_count_tbl <- tibble(
    col = c("Total Unique Patients","Pregnant at Any Entry","Pregnant at Most Recent Entry","Used Pregnant Toggle","Used Pregnant Concerned Toggle"),
  patients = c(
        logfile_count$unique_patients,
        patient_count$pregnant_any,
        patient_count$pregnant_most_recent,
        patient_count$pregnant_toggle,
        patient_count$pregnant_concerned_toggle
  ),
  pct = round((patients/logfile_count$unique_patients) * 100,1)
  
)
kable(pregnant_count_tbl,  col.names = c("Category","Count","%"), align='l',caption="<h1>Patient Pregnancy Status</h1>") %>%
  kable_styling(bootstrap_options = c("striped"), full_width = T) %>%
  pack_rows("Pregnancy Status", 2, 3) %>%
  pack_rows("Toggle Use", 4, 5) 



```

## Toggle Switch Usage

Toggle switches are provided in the Dashboard to allow users to tell the CDS to consider additional information about the patient. For example, if the patient is currently immunosuppressed, but this isn't reflected in the patient record, the "Immunosuppressed" toggle may be switched on, and the CDS will evaluate the patient as if they are immunosuppressed. 

These toggles may be switched on and off as many times as the user likes. Every time a toggle switch is changed, the CDS will automatically re-fire, compute the recommendation again, and add another log entry for the patient. This table summarizes the number of times each toggle switch as set to 'on' across every log entry. It does not show whether the recommendation was changed, only that the toggle was set. 

```{r toggle-switch-usage}
# logdata_messages has all toggle switch usage.
toggle_usage_tbl <- tibble(
  col = c("Total Log Entries","Immunosuppressed","Pregnant","Future pregnancy concerns","Symptomatic","Any Toggle Changed (On or Off)"),
  toggle_counts = c(
    logfile_count$total_entries,
    sum(logdata_messages$isImmunosuppressed),
    sum(logdata_messages$isPregnant),
    sum(logdata_messages$isPregnantConcerned),
    sum(logdata_messages$isSymptomatic),
    sum(logdata_messages$isToggleChanged)
  ),
  pct = round(toggle_counts/logfile_count$total_entries * 100,1)
)
kable(toggle_usage_tbl, col.names = c("","Count","%"), align='l', caption="<h1>Toggle Switch Usage</h1>") %>%
    kable_styling(bootstrap_options = c("striped"), full_width = T) %>%
  pack_rows("Toggle Switch Use", 2, 6)


```

## CDS Performance
The performance of the CDS is dependent on the size of the patient history and how complex the logic is required to produce a recommendation. It is also affected by external factors, such as the load on the FHIR server, database and network latency. It is important to the user experience that the CDS is able to read a patient history and generate a recommendation as quickly as possible. 

```{r cds-performance}
# For now, just generate a histogram of overall applyCds timings. 
# Additional performance metrics can be developed later.
# logdata_messages has the timings for apply

apply_summary <- round(summary(logdata_messages$cdsApplyTime/1000),1)
apply_tbl <- tibble(col=c("Min","1st Quartile","Mean","3rd Quartile","Max"),
                    val=c(apply_summary[1],
                          apply_summary[2],
                          apply_summary[4],
                          apply_summary[5],
                          apply_summary[6])
)
kable(apply_tbl, col.names = c("Value","Time (sec)"), align='l', caption="<h1>CDS Apply Time</h1>") %>%
    kable_styling(bootstrap_options = c("striped"), full_width = T)
```


This is a histogram of the time required by the CDS to generate a recommendation (the "apply" time). It is only one measure of the overall performance of the CDS and Dashboard (for example, it does not account for the time required to fetch the patient information and history from the EHR FHIR Server).

(NOTE: This histogram will look strange until there is a sufficient number of results)

```{r apply-histogram}
# Should consider checking for a minimum number of entries before deciding to draw a histogram

apply_plot <- logdata_messages %>%
  ggplot(aes(x=cdsApplyTime) ) +
    geom_histogram()  +
    geom_density(alpha=.2, fill="#FF6666") +
    ggtitle("CDS Apply Time") +
    xlab("Apply Time in ms")
apply_plot
```

# Recommendation Pathways

The CCSM CDS is broadly structured in two pathways: Screening recommendations and Management recommendations. Screening recommendations apply when the patient is eligible for routine screening, and has **not** had any of the following:

- Recent Abnormal Screening
- Recent Abnormal Histology
- High-grade pre-cancer cervical lesion
- Cervical cancer diagnoses
- High-grade or cancer histology results

Patients may receive specialized recommendations for screening if they are under 30, over 65, currently pregnant, immunocompromised, had exposure to DES in utero or are experiencing abnormal uterine/vaginal bleeding.

The Management pathway generates recommendations when the patient has an abnormal result or pre-cancer diagnoses. Patients may receive specialized recommendations for management if they have 'common' abnormalities (HPV-positive), 'rare' abnormalities (CIN3), or are a special population (under 25, pregnant, immunosuppressed or have had a hysterectomy). There are many recommendations for specific circumstances not covered by the general risk tables. 

```{r recommendation-pathways}
# How many patients received a screening recommendation?
# What pathways/groups did they receive those recommendations from?
# We will look for ScreeningLibrary.DecisionAids.recommendation != null
# We want a table with log entries and the decision aid information for Screening and Management pathways

# Build a table showing all log entries, toggle settings and Management and Screening recommendations
# We want to include the toggle settings so that we can filter recommendations that were made with 
# no toggles set vs. those with toggle switches set. We also add the timings to see how
# performance varies by pathway.
logdata_pathways <-tidyjson::as_tibble(
  logdata %>%
    spread_values(id = jstring(id)) %>% 
    enter_object(message) %>%
      spread_values(age = jinteger('patientInfo','age')) %>%
      enter_object(payload) %>%
        spread_values(
          mgmt_hasRecommendation = jstring('ManagementLibrary','HasRecommendation'),
          mgmt_recommendation = jstring('ManagementLibrary','DecisionAids','recommendation'),
          mgmt_recommendationDate = jstring('ManagementLibrary','DecisionAids','recommendationDate'),
          mgmt_recommendationGroup = jstring('ManagementLibrary','DecisionAids','recommendationGroup')
        )  %>%
        spread_values(
          scrn_hasRecommendation = jstring('ScreeningLibrary','HasRecommendation'),
          scrn_recommendation = jstring('ScreeningLibrary','DecisionAids','recommendation'),
          scrn_recommendationDate = jstring('ScreeningLibrary','DecisionAids','recommendationDate'),
          scrn_recommendationGroup = jstring('ScreeningLibrary','DecisionAids','recommendationGroup')
        ) %>%
      mutate(
        mgmt_recommendationDate = date(mgmt_recommendationDate),
        scrn_recommendationDate = date(scrn_recommendationDate)) %>%
  select(-document.id)  
) %>%
  left_join(logdata_messages, by='id') %>% # Add columns which have toggle status and timings 
  arrange(desc(timeRequestSent))

# all 'default' entries for all patients
logdata_pathways_notoggle <- logdata_pathways %>%
  filter(!isToggleChanged)

rec_counts = list(
  recs_all = 0,
  mgmt_recs_all = 0,
  scrn_recs_all = 0,
  mgmt_recs_def = 0,
  scrn_recs_def = 0
)
rec_counts$recs_all <- nrow(logdata_pathways)
rec_counts$recs_def <- nrow(logdata_pathways_notoggle)

# We need to check that screening recommendation is null; 
# A patient who would normally get a management recommendation 
# will get a screening recommendation if they are symptomatic. 
# They will have both a screening recommendation and a management
# recommendation, but the screening one 'wins'
management_recs_all_tbl <- logdata_pathways %>%
  filter(mgmt_hasRecommendation == TRUE, scrn_hasRecommendation == FALSE) %>%
  select(-scrn_recommendation, -scrn_recommendationDate, -scrn_recommendationGroup)

rec_counts$mgmt_recs_all <- nrow(management_recs_all_tbl)

# get the 'base' recommendations generated when no toggle switches were used
# isToggleChanged is FALSE when all toggles are initially off. 
management_recs_notoggle_tbl <- management_recs_all_tbl %>%
  filter(!isToggleChanged)

rec_counts$mgmt_recs_def <- nrow(management_recs_notoggle_tbl)

# this is still a table of all log entries, just the default no toggle state
# there could be multiple entries for the same patient (reload, refresh, etc)

# Note it is possible to have a screening recommendation AND a management recommendation
# if the patient is symptomatic. 
screening_recs_all_tbl <- logdata_pathways %>%
  filter(mgmt_hasRecommendation == FALSE, scrn_hasRecommendation == TRUE)

screening_recs_notoggle_tbl <- screening_recs_all_tbl %>%
  filter(!isToggleChanged)

rec_counts$scrn_recs_all <- nrow(screening_recs_all_tbl)
rec_counts$scrn_recs_def <- nrow(screening_recs_notoggle_tbl)

rec_counts_tbl <- tibble(
  cols=c("Management Recommendations", "Screening Recommendations","Total","Management Recommendations","Screening Recommendations","Total"),
  count = c(rec_counts$mgmt_recs_all, rec_counts$scrn_recs_all, rec_counts$recs_all, rec_counts$mgmt_recs_def, rec_counts$scrn_recs_def, rec_counts$recs_def),
  pct = round(count/rec_counts$recs_all * 100,1)
)
kable(rec_counts_tbl,  col.names = c("Pathway","Count","% of All Entries"), align='l',caption="<h1>Pathway Recommendation Counts</h1>") %>%
  kable_styling(bootstrap_options = c("striped"), full_width = T) %>%
  pack_rows("All Entries", 1, 3) %>%
  pack_rows("Default Recommendations", 4,6)
```


If the counts for "all entries" are larger than for "default" that means that the toggle switches have been used when viewing a patient in the Dashboard. Switching a toggle switch to "on" and back to "off" will cause two log entries to be generated. "Default" means the recommendation was generated without using any toggle switches, but can still include repeat viewings of the same patient in the Dashboard.

## Screening Pathway

The Screening pathway includes recommendations arranged by special groups. We can summarize which groups produced screening recommendations for patients. These are across all log entries.

```{r screening-rec-groups}

# Summarize recommendations for Screening vs Management
# Note it is possible to have a screening recommendation AND a management recommendation
# if the patient is symptomatic. 

# Count by groups
screening_groups_tbl <- screening_recs_all_tbl %>%
  count(scrn_recommendationGroup)
kable(screening_groups_tbl,  align='l', col.names = c('Group','Count'), caption="<h1>Screening Groups (All log entries)</h1>") %>%
  kable_styling(bootstrap_options = c("striped"), full_width = T)

# Count by recommendation and group
screening_rec_group_tbl <- screening_recs_all_tbl %>% group_by(scrn_recommendation,scrn_recommendationGroup) %>% summarize(Count = n())
kable(screening_rec_group_tbl,  align='l', col.names = c('Recommendation',  'Group','Count'), caption="<h1>Screening Recommendations (All log entries)</h1>") %>%
  kable_styling(bootstrap_options = c("striped"), full_width = T)


```

We can also look at the distinct recommendations patients received when evaluated by the CDS with no toggle switches used. Note that a patient may receive the same screening recommendation multiple times because they are seen on multiple visits, or the CDS Dashboard was opened multiple times for the same patient. 

```{r screening-recs-by-patient}
# get the 'base' recommendation with no toggle switches used
screening_recs_by_patient <- screening_recs_notoggle_tbl %>%
  group_by(patientReference, scrn_recommendation, scrn_recommendationGroup) %>%
  summarize(count = n()) 

screening_recs_by_type <- screening_recs_by_patient %>%
  group_by(scrn_recommendation, scrn_recommendationGroup) %>%
  summarize(patients = n())

kable(screening_recs_by_type,  align='l', col.names = c('Recommendation',  'Group','Patient Count'), caption="<h1>Screening Recommendations (Unique Patients)</h1>") %>%
  kable_styling(bootstrap_options = c("striped"), full_width = T)  
```


## Under 30 Screening

The CDS contains specific recommendations for patients under age 30. We can see the count of patients under 30 who received screening recommendations. Again, note that individual patients may receive more than one recommendation because they have been evaluated by the CDS on more than one occassion. 

```{r under-30-screening}
# We can filter those patients under 30 from the notoggle table
# All counts of a patient under 30 receiving a recommendation/group 
screening_under30_by_patient <- screening_recs_notoggle_tbl %>%
  filter(age < 30) %>%
  group_by(patientReference,scrn_recommendation, scrn_recommendationGroup) %>%
    summarize(count = n())
if (nrow(screening_under30_by_patient) > 0)  {
  
 
  screening_under30_by_type <- screening_under30_by_patient %>%
    group_by(scrn_recommendation, scrn_recommendationGroup) %>%
    summarize(patients = n())
  
  kable(screening_under30_by_type,  align='l', col.names = c('Recommendation',  'Group','Patient Count'), caption="<h1>Under 30 Screening Recommendations (Unique Patients)</h1>") %>%
    kable_styling(bootstrap_options = c("striped"), full_width = T)  
  } else {
  cat("No patients under 30 received a screening recommendation.")
}
```

## Additional Screening Results
Additional screening results will be added here. 



## Management Pathway Recommendations

The Management Pathway covers recommendations for 2019 ASCCP Risk-Based Management Consensus Guidelines, and is organized in three main pathways: Common Abnormalities & Risk Tables, Rare Abnormalities (Guideline sections G, H, I and J) and Special Populations (Under 25, Immunocompromised, Pregnant, Hysterectomy).

```{r management-recommendations}
# We want to determine the management pathway which produced each recommendation
# If ManageSpecialPopulation.WhichPopulationMadeTheRecommendation is not null then Special Population
# If ManageRareAbnormality.WhichRarityMadeTheRecommendation is not null then Rare Abnormality
# else Common Abnormality

# We need to parse out the log data carefully, as there are values/objects which can be null
# causing an error if we simply try to spread_all across the entire payload.
# Note that this pass does not expand all nested objects - we'll do that as needed for
# individual metrics

management_data_dashboard <-tidyjson::as_tibble(
  logdata %>%
    spread_values(id = jstring(id)) %>% 
    enter_object(message) %>%
      enter_object(payload,DashboardLibrary) %>%
        spread_all
)

management_data_library <-tidyjson::as_tibble(
  logdata %>%
    spread_values(id = jstring(id)) %>% 
    enter_object(message) %>%
      enter_object(payload,ManagementLibrary) %>%
        spread_all(recursive = FALSE)
)

management_data_common <-tidyjson::as_tibble(
  logdata %>%
    spread_values(id = jstring(id)) %>% 
    enter_object(message) %>%
      enter_object(payload,ManageCommonAbnormality) %>%
        spread_all(recursive = FALSE) %>%
        mutate(DateOfMostRecentHpvReport = date(DateOfMostRecentHpvReport),
               DateOfMostRecentPositiveScreeningResult = date(DateOfMostRecentPositiveScreeningResult))
)

# In the Risk Tables, there are management recommendations which are not associated with specific risk estimates.
# For coding purposes, these are represented as -1 in the lookup tables used by the CDS. We'll need to replace these with "NA"
management_data_relevant_row <-tidyjson::as_tibble(
  logdata %>%
    spread_values(id = jstring(id)) %>% 
    enter_object(message) %>%
      enter_object(payload,ManageCommonAbnormality,RelevantRow) %>%
        spread_all(recursive = FALSE) %>%
        mutate(riskNow = na_if(riskNow, -1), risk5yr = na_if(risk5yr, -1))
)

management_data_rare <-tidyjson::as_tibble(
  logdata %>%
    spread_values(id = jstring(id)) %>% 
    enter_object(message) %>%
      enter_object(payload,ManageRareAbnormality) %>%
        spread_all(recursive = FALSE)
)

management_data_special <-tidyjson::as_tibble(
  logdata %>%
    spread_values(id = jstring(id)) %>% 
    enter_object(message) %>%
      enter_object(payload,ManageSpecialPopulation) %>%
        spread_all(recursive = FALSE)
)
management_data_collate <-tidyjson::as_tibble(
  logdata %>%
    spread_values(id = jstring(id)) %>% 
    enter_object(message) %>%
      enter_object(payload,CollateManagementData) %>%
        spread_all(recursive = FALSE) %>%
        mutate(MostRecentBiopsyReportDate = date(MostRecentBiopsyReportDate),
               MostRecentCytologyReportDate = date(MostRecentCytologyReportDate),
               DateOfFirstBiopsyAfterMostRecentHpvReport = date(DateOfFirstBiopsyAfterMostRecentHpvReport),
               DateOfMostRecentCytologyBeforeBiopsy = date(DateOfMostRecentCytologyBeforeBiopsy))
)
# Join 'em up!
management_data_all <- management_recs_all_tbl %>%
  left_join(management_data_dashboard) %>%
  left_join(management_data_common) %>%
  left_join(management_data_rare) %>%
  left_join(management_data_special) %>%
  mutate(Pathway = ifelse(!is.na(WhichPopulationMadeTheRecommendation),"Special Population",
                          ifelse(!is.na(WhichRarityMadeTheRecommendation),"Rare Abnormality",
                          "Common Abnormality")))

management_pathway_recs <- management_data_all %>%
  select(id,Pathway, Group = mgmt_recommendationGroup, Recommendation = mgmt_recommendation)

management_pathway_plot <- management_pathway_recs %>% ggplot(aes(x = Pathway, fill=Recommendation, )) +
  geom_bar(position="stack") +
  ggtitle("Recommendation Count by Management Pathway") +
  xlab("Management Pathway") + 
  ylab("Recommendation Count")
management_pathway_plot
```
This chart shows the distribution of recommendations by the three main management pathways. Individual breakdowns by specific pathways will follow. 

```{r}
management_pathway_group_tbl <- management_pathway_recs %>% group_by(Pathway,Group) %>% summarize(Count = n())

kable(management_pathway_group_tbl,  align='l', col.names = c('Pathway',  'Group','Recommendation Count'),caption="<h1>Management Recommendations Summary</h1>") %>%
  kable_styling(bootstrap_options = c("striped"), full_width = T)

```

### Management Recommendation Groups
There are many ways for a patient to receive the same recommendation in the Management pathways. These are organized into groups of related guideline logic. This table shows the breakdown of different recommendations by guideline groups. 

```{r}
management_rec_group_tbl <- management_pathway_recs %>%
  group_by(Recommendation,Group) %>%
  summarize(Count = n()) %>% 
  arrange(Recommendation, desc(Count))

kable(management_rec_group_tbl,  align='l', col.names = c('Recommendation',  'Group','Count'), caption="<h1>Management Recommendation Counts by Group</h1>") %>%
  kable_styling(bootstrap_options = c("striped"), full_width = T)
```

## Special Population Recommendations

The CDS pathway for special populations include guideline recommendations for several different groups of patients. These recommendations are given priority over other possible recommendations.

```{r special-populations}
# What is the count of recommendations by population group?
special_population_groups <- c("Patients with Histologic Cancer","Patients Younger than 25","Pregnant Patients","Immunosuppressed Patients","Managing Patients After Hysterectomy")
# Get the list of special population recommendations non-toggle. This is just the count of how many times the CDS activated a special population recommendation. Can this be filtered
# to the most recent log entry per patient per day?
# Then add in the immunosuppressed and pregnant toggle recommendations. This is just the count of how many times the CDS triggered the recommendation based on the toggle.
# Then just filter to the Most Recent unique patient entry. This is the 'current' recommendation from special population.
# Show by WhichPopulationMadeTheRecommendation with sub-group counts.
```



## Risk Table Recommendations

The recommendations generated in the Common Abnormality pathway includes those produced by the Risk Tables. We can count the times a recommendation was produced from the Risk Tables. This is a specific subset of Management guidelines, and does not represent the full scope of possible recommendations. 

```{r}
# Table 3 Row lookup does not return referring HPV and Cytology as result, so we coalesce columns for this.
# Note management patients receiving a management recommendation should receieve a valid
# risk table recommendation.
# Select distinct log entries for a patientReference, recommendation date, recommendation and group. 
# Filter out hysterectomy and immunosuppressed recommendations which supercede risk tables
management_risk_table_recs <- management_data_all |>
  filter(!is.na(WhichTableMadeTheRecommendation), mgmt_recommendation != "Hysterectomy", mgmt_recommendationGroup != "Immunosuppressed (K.3)") |>
  distinct(patientReference,mgmt_recommendationDate,mgmt_recommendation,mgmt_recommendationGroup, .keep_all = TRUE) %>%
  left_join(management_data_relevant_row) %>%
  left_join(management_data_collate) %>%
  mutate(risk_table_cyto = ifelse(cyto=="",ReferringCytologyResult,cyto)) %>%
  mutate(risk_table_hpv = ifelse(hpv=="",ReferringHpvResult,hpv))


management_risk_table_recs_plot <- management_risk_table_recs %>% ggplot(aes(x=TableRecommendation)) +
  ggtitle("Risk Table Recommendations") +
  ylab("Count of Recommendations") +
  xlab("Risk Table Recommendation") +
  geom_bar()

management_risk_table_recs_plot
```

```{r risk-tables-table}
management_risk_group_tbl <- management_risk_table_recs %>%
  group_by(TableRecommendation,mgmt_recommendationGroup) %>%
  summarize(count = n()) %>% 
  mutate(percent = percent(count / nrow(management_risk_table_recs))) %>%
  arrange(TableRecommendation, desc(count))

management_risk_group_tbl <- janitor::adorn_totals(management_risk_group_tbl, where="row")

kable(management_risk_group_tbl,  align='l', col.names = c('Recommendation','Group','Count','Percent'), caption="<h1>Risk Table Recommendation Counts</h1>") %>%
  kable_styling(bootstrap_options = c("striped"), full_width = T)
```

# Most Recent Cytology and HPV for Risk Table Recommendations
The risk table-based recommendations are based on current HPV and Cytology results. 
For patients who received a recommendation from the risk tables, what was their most recent cytology and HPV results?

```{r}
risk_table_cyto <- management_risk_table_recs %>%
#  ungroup() %>%
  count(cyto, name="count") %>%
  mutate(percent = percent(count / nrow(management_risk_table_recs))) %>%
  rename(result = cyto)
risk_table_cyto[risk_table_cyto$result == "ALL",1] <- "No Result"
risk_table_cyto[risk_table_cyto$result == "",1] <- "No Result"


risk_table_hpv <- management_risk_table_recs %>%
  # ungroup() %>%
  count(hpv, name="count") %>%
  mutate(percent = percent(count / nrow(management_risk_table_recs))) %>%
  rename(result = hpv)
risk_table_hpv[risk_table_hpv$result == "ALL",1] <- "No Result"
risk_table_hpv[risk_table_hpv$result == "",1] <- "No Result"
all_risk_table <- tibble(
  result = c("All Results"),
  count = nrow(management_risk_table_recs),
  percent = percent(1)
)
risk_table_cyto_hpv_summary <-  bind_rows(all_risk_table, risk_table_cyto, risk_table_hpv)
kable(risk_table_cyto_hpv_summary,  col.names = c("Result","Count","%"), align='l', caption="<h1>Current Cytology and HPV Results</h1>") %>%
  kable_styling(bootstrap_options = c("striped"), full_width = T) %>%
  pack_rows("All Results", 1, 1) %>%
  group_rows("By Cytology Result", 2, 1+nrow(risk_table_cyto)) %>%
  group_rows("By HPV Result", 2+nrow(risk_table_cyto), 1+nrow(risk_table_cyto)+nrow(risk_table_hpv))

```

# Most Recent Biopsy for Risk Table Recommendations
Risk Table recommendations are also computed based on the patient's most recent biopsy, or a history of having been treated for CIN2 or CIN3. 
This summarizes the most recent and second most recent biopsy results for patients who received a risk table recommendation.

```{r biopsy-table-3}
risk_table_most_recent_biopsy_count <- management_risk_table_recs %>%
  count(MostRecentBiopsyResult, name="count") %>%
  mutate(percent = percent(count / nrow(management_risk_table_recs))) %>%
  rename(result = MostRecentBiopsyResult)
risk_table_most_recent_biopsy_count[is.na(risk_table_most_recent_biopsy_count$result),1] <- "No Result"
risk_table_most_recent_biopsy_count[risk_table_most_recent_biopsy_count$result == "UNK",1] <- "Unknown"


risk_table_second_most_recent_biopsy_count <- management_risk_table_recs %>%
  count(SecondMostRecentBiopsyResult, name="count") %>%
  mutate(percent = percent(count / nrow(management_risk_table_recs))) %>%
  rename(result = SecondMostRecentBiopsyResult)
risk_table_second_most_recent_biopsy_count[is.na(risk_table_second_most_recent_biopsy_count$result),1] <- "No Result"
risk_table_second_most_recent_biopsy_count[risk_table_second_most_recent_biopsy_count$result == "UNK",1] <- "Unknown"
all_risk_table <- tibble(
  result = c("All Results"),
  count = nrow(management_risk_table_recs),
  percent = percent(1)
)
risk_table_biopsy_summary <-  bind_rows(all_risk_table,risk_table_most_recent_biopsy_count,risk_table_second_most_recent_biopsy_count)
kable(risk_table_biopsy_summary,  col.names = c("Result","Count","%"), align='l', caption="<h1>Biopsy Results</h1>") %>%
  kable_styling(bootstrap_options = c("striped"), full_width = T) %>%
  group_rows("Most Recent Biopsy Result", 2, 1+nrow(risk_table_most_recent_biopsy_count)) %>%
  group_rows("Second Most Recent Biopsy Result", 2+nrow(risk_table_most_recent_biopsy_count), 1+nrow(risk_table_most_recent_biopsy_count)+nrow(risk_table_second_most_recent_biopsy_count))
```

# Risk Table Recommendations by Immediate CIN3+ Risk
Risk Table recommendations are based on the estimated CIN3+ risk. If the risk is 4% or greater, immediate management via colposcopy or treatment is indicated. If the immediate risk is less
than 4%, the 5-year CIN 3+ risk is examined to determine whether patients should return in 1, 3, or 5 years.
If the immedaite risk for CIN 3+ is 60% or greater, expedited treatment without confirmatory biopsy is recommended.

```{r cin3-risk-immediate}
risk_table_risk_summary <- management_risk_table_recs %>% filter(WhichTableMadeTheRecommendation != 3) %>%
  group_by(action, riskNow, risk5yr) %>% count() %>%
  mutate(percent = percent(n / nrow(management_risk_table_recs)))

kable(risk_table_risk_summary,  col.names = c("Recommendation","CIN 3+ Immediate Risk %", "CIN 3+ 5yr Risk %", "Count","%"), align='l', caption="<h1>CIN 3+ Risks by Recommended Action</h1>") %>%
  kable_styling(bootstrap_options = c("striped"), full_width = T) 

```

# Risk Estimates for Colposcopy/Biopsy Result
The evaluation of a biopsy result is made using the recommended management action from Table 3 of the Risk Estimates for Consensus Guidelines.
Current biopsy results of CIN2 or greater are not associated with specific risk estimates.
```{r cin3-risk-biopsy}

risk_table_biopsy_risk_summary <- management_risk_table_recs %>%
  filter(WhichTableMadeTheRecommendation == 3) %>%
  group_by(action, riskNow, risk5yr) %>%
  count(name="count") %>%
  ungroup() %>%
  mutate(percent = percent(count / sum(count), accuracy = 0.1))
# risk_table_biopsy_risk_summary

kable(risk_table_biopsy_risk_summary,  col.names = c("Recommendation","CIN 3+ 1yr Risk %", "CIN 3+ 5yr Risk %", "Count","%"), align='l', caption="<h1>CIN 3+ Risks by Recommended Action for Biopsy Results</h1>") %>%
  kable_styling(bootstrap_options = c("striped"), full_width = T) 

```
# Expedited Treatment
What are the histology, HPV and cytology results associated with patients who received a recommendation for expedited treatment?

```{r expedited-treatment}
# Table 3 recommendations for expedited treatment for CIN2, CIN3, AIS, Cancer

expedited_treatment_table3 <- management_risk_table_recs %>%
  filter(action == "Treatment", WhichTableMadeTheRecommendation == 3) %>%
  select(id,patientReference,timeRequestSent,mgmt_recommendation,mgmt_recommendationGroup,WhichTableMadeTheRecommendation,TableRecommendation,action,risk_table_cyto,risk_table_hpv,biopsy,riskNow,risk5yr) %>%
  group_by(result = biopsy,riskNow,risk5yr) %>%
  summarize(count = n()) %>% ungroup()
# Table 1 recommendation for expedited treatment for HPV16+/HSIL+
expedited_treatment_table1 <- management_risk_table_recs %>%
  filter(action == "Treatment", WhichTableMadeTheRecommendation == 1) %>%
  select(id,patientReference,timeRequestSent,mgmt_recommendation,mgmt_recommendationGroup,WhichTableMadeTheRecommendation,TableRecommendation,action,risk_table_cyto,risk_table_hpv,biopsy,riskNow,risk5yr) %>%
  unite(result, risk_table_hpv, risk_table_cyto, sep = "/", remove = FALSE) %>%
  group_by(result,riskNow,risk5yr) %>%
  summarize(count = n()) %>% ungroup()

expedited_treatment_summary <- rbind(expedited_treatment_table3, expedited_treatment_table1) %>%
  mutate(percent = percent(count / sum(count), accuracy = 0.1))

kable(expedited_treatment_summary,  col.names = c("Result", "CIN 3+ Immediate Risk %", "CIN 3+ 5yr Risk %","Count","%"), align='l', caption="<h1>Diagnostic Results for Expedited Treatment</h1>") %>%
  kable_styling(bootstrap_options = c("striped"), full_width = T) 
```

```{r}
# Here's how to drill into an array 
management_data_any_cyto_ascus <-tidyjson::as_tibble(
  logdata %>%
    spread_values(id = jstring(id)) %>% 
    enter_object(message) %>%
      enter_object(payload,CollateManagementData) %>%
        # spread_all(recursive = FALSE) %>%
        enter_object(AnyCytologyInterpretedAsAscusOrAbove) %>%
        gather_array %>%
        spread_all
)

biopsy_dates <- as_tibble (
  logdata %>%
    spread_values(id = jstring(id)) %>%
      enter_object(message,payload,CollateManagementData) %>%
        enter_object(BiopsyDates) %>%
          gather_array() %>% append_values_string(column.name = "BiopsyDate")

)
```

## HPV and Cytology

```{r hpv-cytology-crosstab}
# The intent is to compare HPV results with Cytology results
# We need to look at all patient default recommendations, because they may be seen on different dates
# with different results. HPV-/NILM patients with no abnormalities in recent history would get a screening recommendation
# but the main point of interest will be those with abnormalities who receive a management recommendation.

# Get unique patient default entries (no toggle switches applied)
# And add collate data (which includes MostRecentCytologyCotestResult and MostRecentHpvResult)
cotest_cols <- c("Age Group" = "age_group","Most Recent Cytology Cotest Result" = "MostRecentCytologyCotestResult","HPV-" = "HPV-negative","HPV+" = "HPV-positive","HPV16+" = "HPV16+","HPV16-/18+" = "HPV16-, HPV18+","Unknown"="UNK","N/A"="<NA>")
cotest_results_tbl <- logdata_pathways_notoggle %>%
  left_join(management_data_collate) %>% 
  select(id,patientReference,age,MostRecentCytologyCotestResult,MostRecentHpvResult) %>%
  mutate(age_group = cut(age, breaks=c(0,19,24,29,65,120), labels=c("Under 20","20 to 24","25 to 29","30 to 65","Over 65")))


# cotest_results_pivot_tbl <- cotest_results_tbl %>%
#  pivot_wider(names_from = MostRecentHpvResult, values_from = MostRecentCytologyCotestResult, ) %>%
#  janitor::clean_names()

cotest_results_age_group <- cotest_results_tbl %>%
  group_by(age_group,MostRecentCytologyCotestResult, MostRecentHpvResult) %>%
  tally() %>%
  spread(MostRecentHpvResult, n)

cotest_results_age_group <- cotest_results_age_group %>%
  mutate_at(c(3:ncol(cotest_results_age_group)), ~replace(., is.na(.), 0))  

kable(cotest_results_age_group %>% rename(any_of(cotest_cols)),
    align='l',
    caption="<h1>Most Recent Cotest Results by Age Group</h1>"
  ) %>%
  kable_styling(bootstrap_options = c("striped"), full_width = T)

```
    
## Days to Biopsy
What is the mean number of days between a patient's most recent HPV, Cytology or cotest and the first biopsy (histology) result after that?

This is a complex metric, as there are many possible reasons for a biopsy to occur, and patient histories in the real world are often 'irregular' and may include gaps and missing tests. A patient may have a cotest result of "HPV-negative / NILM" and still have a biopsy performed for other reasons. 
This table will be based on the following conditions:

- A patient has received a biopsy after an HPV, Cytology (Pap) or cotest result. 
- The biopsy occurs after the most recent HPV, Cytology or cotest result.
- If there are multiple biopsies after the most recent HPV, Cytology or cotest result, we will look at the first one (oldest) that occurred after the HPV, Cytology or cotest result.
- This only looks at the most recent test-biopsy pair - it does not compute older intervals for past biopsies. 
- The biopsy must have occurred within the BiopsyReferralPeriod, currently within 1 year after the test result.
- It isn't measuring the time of the colposcopic procedure, only the time of the next immediate histology result. 

This table does not consider the result of the biopsy, or what may follow it. It is simply trying to count the days between the most recent test and the biopsy which followed.

```{r days-to-biopsy}
# Table 1 - Biopsy AFTER most recent cotest

biopsy_days_after_referring_test <- logdata_pathways_notoggle %>%
  left_join(management_data_collate) %>%
  left_join(management_data_common)  %>%
  filter(MostRecentBiopsyReportDate > MostRecentCytologyReportDate | MostRecentBiopsyReportDate > DateOfMostRecentHpvReport) %>%
  distinct(MostRecentBiopsyReportDate, .keep_all = TRUE) %>%
  filter(!(ReferringCytologyResult == "ALL" & ReferringHpvResult == "ALL")) %>%
  select(id,patientReference,timeRequestSent,ReferringCytologyResult,ReferringHpvResult,MostRecentBiopsyReportDate,MostRecentCytologyReportDate,DateOfMostRecentCytologyBeforeBiopsy,DateOfMostRecentHpvReport,DateOfFirstBiopsyAfterMostRecentHpvReport) %>%
  mutate(daysToBiopsy = as.numeric(difftime(coalesce(DateOfFirstBiopsyAfterMostRecentHpvReport,MostRecentBiopsyReportDate), coalesce(DateOfMostRecentCytologyBeforeBiopsy,DateOfMostRecentHpvReport), units = c("days"))))

biopsy_days_cyto_summary <- biopsy_days_after_referring_test %>%
  group_by(ReferringCytologyResult) %>%
  summarize(days = round2str(mean(daysToBiopsy)), count = n(), percent = percent(count/nrow(biopsy_days_after_referring_test), accuracy = 0.1)) %>%
  rename("Result" = "ReferringCytologyResult") %>%
  arrange(factor(Result, levels = c("ALL","NILM","ASC-US","LSIL","ASC-H","AGC","HSIL+")))
biopsy_days_cyto_summary[biopsy_days_cyto_summary$Result == "ALL",1] <- "Unknown/Missing"

biopsy_days_hpv_summary <- biopsy_days_after_referring_test %>%
  group_by(ReferringHpvResult) %>%
  summarize(days = round2str(mean(daysToBiopsy)), count = n(), percent = percent(count/nrow(biopsy_days_after_referring_test), accuracy = 0.1)) %>%
  rename("Result" = "ReferringHpvResult")
biopsy_days_hpv_summary[biopsy_days_hpv_summary$Result == "ALL",1] <- "Unknown/Missing"

mean_all_biopsy_days <- tibble(
  Result = c("All Results"),
  days = c(round2str(mean(biopsy_days_after_referring_test$daysToBiopsy))),
  count = nrow(biopsy_days_after_referring_test),
  percent = percent(1)
)


biopsy_days_summary <-  bind_rows(mean_all_biopsy_days,biopsy_days_cyto_summary, biopsy_days_hpv_summary)

kable(biopsy_days_summary,  col.names = c("Result","Days","Count","%"), align='l',caption="<h1>Mean Days to Biopsy</h1>") %>%
  kable_styling(bootstrap_options = c("striped"), full_width = T) %>%
  pack_rows("All Results", 1, 1) %>%
  group_rows("By Cytology Result", 2, 1+nrow(biopsy_days_cyto_summary)) %>%
  group_rows("By HPV Result", 2+nrow(biopsy_days_cyto_summary), 1+nrow(biopsy_days_cyto_summary)+nrow(biopsy_days_hpv_summary))
```

# 